A boundary-integral framework to simulate \\ viscous erosion of a porous medium

### 2018-04-14 docs/methods/erosion_methods.tex
NM

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction\label{s:intro}}
Flow-induced erosion deteriorates and reshapes solid material over a range of scales found in nature, from large land formations~\cite{han1969, perron2009formation, abrams2009growth, petroff2011geometry, jerolmack2012internal, Rothman2012, coh-dev-sey-yi-szy-rot2015, perkins2015amplification}, to centimeter-scale features and patterns~\cite{daerr2003erosion, berhanu2012shape, nienhuis2014wavelength, domokos2014river}. Though less visible, these same forces are working at the very smallest scales, slowly deteriorating the individual constituents of porous media (e.g.~soil, sand, or clay) or biological structures like plaque and biofilms~\cite{pic-van-hei2000, sha2002, gro-gij-van-fer-hat-van-yua-wen2007}. Motivated by such examples, this paper develops numerical methods to simulate fluid-mechanical erosion of multiple solid bodies in a Stokes flow---the most relevant regime for groundwater \cite{bear2013dynamics, cao2010coupled} and the aforementioned bio-fluid applications.

Recent work in the high-Reynolds-number regime has highlighted the importance of a shape-flow feedback that occurs during erosion~\cite{ris-moo-chi-she-zha2012, moo-ris-chi-zha-she2013, hewett2017evolution, moore2017riemann, lopez2018cfd} and during related processes such as dissolution and melting~\cite{Huang2015, kondratiuk2015steadily, dallaston2015channelization, rycroft2016asymmetric, cohen2016erosion, hewett2017pear, claudin2017dissolution}.  In these cases, the fluid alters the morphology of the immersed structures, which in turn modifies the surrounding flow field, and so on. Such examples of `sculpting with flow'~\cite{ristroph2018sculpting} have been examined in the context of a single body~\cite{ris-moo-chi-she-zha2012, moore2017riemann} and, more recently, multiple bodies whose shapes coevolve as mediated by the flow~\cite{hewett2017evolution}. From this viewpoint, the example of an eroding porous medium presents new opportunities and challenges. First, since the medium is composed of many individual constituents, multiple bodies must be considered. Second, the dense packing present in typical porous-media applications requires close-range interactions to be resolved. Finally, long-range effects inherent to Stokes flow couples the shape evolution of {\em all} bodies to a greater degree than any other flow regime.

@@

At the next stage, we must use this shear stress to advance solid boundaries. Tracking fluid-structure interfaces with long time horizons presents numerical challenges in that the underlying mesh can become distorted and tangled. Furthermore, the process of erosion naturally leads to regions of high curvature and even corners.  We address these tracking issues in two ways: first, we introduce a tangential velocity that keeps the grid equispaced with respect to arc length but that does not affect dynamics; second, we suppress high-frequency oscillations in the interface evolution via a curvature-penalization. The interface is parametrized in the so-called {\thL} coordinate system so that the curvature penalization appears as a diffusive term \cite{hou-low-she1994}. This term can be easily be handled in a numerically stable fashion with an exponential integrator.  Once the simulation is complete, we perform a post-processing step in which we compute bulk properties, such as vorticity and pressure, using an upsampled trapezoid rule.

We supplement our primarily computational study with a few analytical results. In particular, our simulation shows that a single eroding body tends to a slender, for-aft symmetric form characterized by nearly uniform shear stress along its surface. This body features sharp corners at its front and rear stagnation points. The principle of long-time uniform stress combined with some local analysis leads to an analytical prediction for the opening angles formed at the front and rear. Additionally, drag laws from slender-body theory allows us to derive scaling laws for the rate of the body's area and drag reduction. These analytical results, in addition to their own intrinsic value, allow us to conclusively validate the numerical simulation.

\paragraph{Limitations}
We make two fundamental design choices in our simulation: first, we restrict attention to two spatial dimensions, and second the bodies are considered immobile (i.e.~fixed to their initial locations). The restriction to two dimensions allows us to simulate the erosion of many bodies (up to $50$ in this manuscript) with high resolution, enabling accurate characterization of porous-medium erosion. While BIEs are available in three dimensions, the extra computational cost limits the number of bodies that could be practically simulated and makes porous-medium simulations less feasible. Second, the assumption of immobile bodies affords a simple framework to analyze the effects of erosion in isolation. While other mechanisms, such as sedimentation or transport, undoubtedly play a role in real applications, these effects would introduce additional complexity, both conceptually and numerically. We therefore defer the incorporation of these mechanisms to future work.

\paragraph{Related Work}
In a closely related study, Mitchell \& Spagnolie used a traction integral equation formulation to simulate viscous erosion of a single body in three dimensions~\cite{mit-spa2016}. The flexibility of this formulation enabled simulations of a {\em mobile} body eroding in an {\em aribtrary} background flow (e.g.~uniform flow, shear flow, etc.).  An advantage of the traction formulation is that is directly provides the shear stress that is needed to evolve surfaces.  In the present work, we aim to simulate the erosion of a porous medium and, further, choose to fix the bodies to their initial locations. Thus, rather than a {\em single, mobile} body as considered in~\cite{mit-spa2016}, we aim to simulate the erosion of {\em many, immobile} bodies. Although the traction formulation has several attractive features, we instead employ a standard double-layer formulation since it allows us to easily calculate bulk quantities (e.g.~vorticity, pressure, and velocity) that will assist our understanding of the erosion process.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction\label{s:intro}}
Flow-induced erosion deteriorates and reshapes solid material over a range of scales found in nature, from massive land formations~\cite{han1969, perron2009formation, abrams2009growth, petroff2011geometry, jerolmack2012internal, Rothman2012, coh-dev-sey-yi-szy-rot2015, perkins2015amplification}, to centimeter-scale features and patterns~\cite{daerr2003erosion, berhanu2012shape, nienhuis2014wavelength, domokos2014river}. Though less visible, these same forces are working at the very smallest scales, slowly deteriorating the individual constituents of porous media (e.g.~soil, sand, or clay) or biological structures like plaque and biofilms~\cite{pic-van-hei2000, sha2002, gro-gij-van-fer-hat-van-yua-wen2007}. Motivated by such examples, this paper develops numerical methods to simulate fluid-mechanical erosion of multiple solid bodies in a Stokes flow---the most relevant regime for groundwater \cite{bear2013dynamics, cao2010coupled} and the aforementioned bio-fluid applications.

Recent work in the high-Reynolds-number regime has highlighted the importance of a shape-flow feedback that occurs during erosion~\cite{ris-moo-chi-she-zha2012, moo-ris-chi-zha-she2013, hewett2017evolution, moore2017riemann, lopez2018cfd} and the related processes of dissolution and melting~\cite{Huang2015, kondratiuk2015steadily, dallaston2015channelization, rycroft2016asymmetric, cohen2016erosion, hewett2017pear, claudin2017dissolution}.  In these cases, the fluid alters the morphology of immersed structures, which in turn modifies the surrounding flow field, and so on. Such examples of `sculpting with flow'~\cite{ristroph2018sculpting} have been examined in the context of a single body~\cite{ris-moo-chi-she-zha2012, moore2017riemann} and, more recently, multiple bodies whose shape evolution is intertwined~\cite{hewett2017evolution}. From this viewpoint, the example of an eroding porous medium presents new opportunities and challenges. First, since the medium is composed of many individual constituents, multiple bodies must be considered. Second, the dense packing present in typical porous-media applications requires close-range interactions to be resolved. Finally, long-range effects inherent to the Stokes limit couples the shape evolution of {\em all} bodies to a greater degree than any other regime.

@@

At the next stage, we must use the shear stress to advance solid boundaries. Tracking fluid-structure interfaces with long time horizons presents numerical challenges in that the underlying mesh can become distorted and tangled. Furthermore, the process of erosion naturally leads to regions of high curvature and even corners.  We address these tracking issues in two ways: first, we introduce a tangential velocity that keeps the grid equispaced with respect to arc length but that does not affect dynamics; second, we suppress high-frequency oscillations in the interface evolution via curvature penalization. The interface is parametrized in the so-called {\thL} coordinate system so that the curvature penalization appears as a diffusive term \cite{hou-low-she1994}. This term can be easily be handled in a numerically stable fashion with an exponential integrator.  Once the simulation is complete, we perform a post-processing step in which we compute bulk properties, such as vorticity and pressure, using an upsampled trapezoid rule.

We supplement our primarily computational study with a few analytical results. In particular, our simulation shows that a single eroding body tends to a slender, for-aft symmetric form characterized by nearly uniform shear stress along its surface. This body exhibits sharp corners at its front and rear. The principle of long-time uniform stress combined with some local analysis leads to an analytical prediction for the opening angles formed at the front and rear. Additionally, slender-body theory gives rise to scaling laws for the reduction rate of the body's area and drag. These analytical results, in addition to their own intrinsic value, allow us to conclusively validate the numerical simulation.

\paragraph{Limitations}
We make two fundamental design choices in the simulations described here: first, we restrict attention to two spatial dimensions, and second the bodies are considered immobile (i.e.~fixed to their initial locations). By restricting to two dimensions, we can simulate many bodies (up to $50$ in this manuscript) with high resolution, enabling accurate characterization of porous-medium erosion. While BIEs are available in three dimensions, the extra computational cost limits the number of bodies that could be practically simulated and makes porous-medium simulations less feasible. Meanwhile, the second assumption of immobile bodies affords a simple framework to explore the effects of erosion in isolation. While other mechanisms, such as sedimentation or transport, undoubtedly play a role in real applications, these effects would introduce additional complexity, both conceptually and numerically. We therefore defer their study to future work.

\paragraph{Related Work}
In a closely related study, Mitchell \& Spagnolie used a traction integral equation formulation to simulate the viscous erosion of a single body in three dimensions~\cite{mit-spa2016}. The flexibility of this formulation enabled simulations of a {\em mobile} body eroding in an {\em aribtrary} background flow (e.g.~uniform flow, shear flow, etc.).  An advantage of the traction formulation is that is directly provides the shear stress needed to evolve surfaces.  In the present work, we aim to simulate the erosion of a porous medium and, further, choose to fix the bodies to their initial locations. Thus, rather than a {\em single, mobile} body as considered in~\cite{mit-spa2016}, we aim to simulate the erosion of {\em many, immobile} bodies. Although the traction formulation has several attractive features, we instead employ a standard double-layer formulation since it allows us to easily calculate bulk quantities (e.g.~vorticity, pressure, and velocity) that will assist our understanding of the erosion process.
